---
title: "Adams_FinalProject"
author: "Grace Adams"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A handful of fyi stuff: 

- cbb.csv has seasons 2013-2019 and seasons 2021-2023 combined

- The 2020 season's data set is kept separate from the other seasons, because there was no postseason due to the Coronavirus.

- cbb24 contains data as of 3/18/2024. This dataset will be updated and added to cbb.csv at the conclusion of the Tournament.

Data Dictionary: 

- RK (Only in cbb20): The ranking of the team at the end of the regular season according to barttorvik

- TEAM: The Division I college basketball school

- CONF: The Athletic Conference in which the school participates in (A10 = Atlantic 10, ACC = Atlantic Coast Conference, AE = America East, Amer = American, ASun = ASUN, B10 = Big Ten, B12 = Big 12, BE = Big East, BSky = Big Sky, BSth = Big South, BW = Big West, CAA = Colonial Athletic Association, CUSA = Conference USA, Horz = Horizon League, Ivy = Ivy League, MAAC = Metro Atlantic Athletic Conference, MAC = Mid-American Conference, MEAC = Mid-Eastern Athletic Conference, MVC = Missouri Valley Conference, MWC = Mountain West, NEC = Northeast Conference, OVC = Ohio Valley Conference, P12 = Pac-12, Pat = Patriot League, SB = Sun Belt, SC = Southern Conference, SEC = South Eastern Conference, Slnd = Southland Conference, Sum = Summit League, SWAC = Southwestern Athletic Conference, WAC = Western Athletic Conference, WCC = West Coast Conference)

- G: Number of games played

- W: Number of games won

- ADJOE: Adjusted Offensive Efficiency (An estimate of the offensive efficiency (points scored per 100 possessions) a team would have against the average Division I defense)

- ADJDE: Adjusted Defensive Efficiency (An estimate of the defensive efficiency (points allowed per 100 possessions) a team would have against the average Division I offense)

- BARTHAG: Power Rating (Chance of beating an average Division I team)

- EFG_O: Effective Field Goal Percentage Shot

- EFG_D: Effective Field Goal Percentage Allowed

- TOR: Turnover Percentage Allowed (Turnover Rate)

- TORD: Turnover Percentage Committed (Steal Rate)

- ORB: Offensive Rebound Rate

- DRB: Offensive Rebound Rate Allowed

- FTR : Free Throw Rate (How often the given team shoots Free Throws)

- FTRD: Free Throw Rate Allowed

- 2P_O: Two-Point Shooting Percentage

- 2P_D: Two-Point Shooting Percentage Allowed

- 3P_O: Three-Point Shooting Percentage

- 3P_D: Three-Point Shooting Percentage Allowed

- ADJ_T: Adjusted Tempo (An estimate of the tempo (possessions per 40 minutes) a team would have against the team that wants to play at an average Division I tempo)

- WAB: Wins Above Bubble (The bubble refers to the cut off between making the NCAA March Madness Tournament and not making it)

- POSTSEASON: Round where the given team was eliminated or where their season ended (R68 = First Four, R64 = Round of 64, R32 = Round of 32, S16 = Sweet Sixteen, E8 = Elite Eight, F4 = Final Four, 2ND = Runner-up, Champion = Winner of the NCAA March Madness Tournament for that given year)

- SEED: Seed in the NCAA March Madness Tournament

- YEAR: Season



## Loading packages

```{r}
#install.packages("tidyverse")
library(tidyverse)
```


## Reading in the data


# Set the directory where your .csv files are located
```{r}
folder_path <- "data_raw/cbb"
```

# List all .csv files in the folder
```{r}
file_list <- list.files(folder_path, pattern = "cbb[0-9]{2}\\.csv")
```


## Loop through each file and read it into the environment
## I don't think this was taught in class, need to find a way to do this with the things taught in class??
```{r}
for (file in file_list) {
  # Extract the two-digit number from the file name
  file_number <- gsub("cbb|\\.csv", "", file)
  
  # Read the .csv file into a data frame with a name based on the file number
  assign(paste0("cbb", file_number), read.csv(file.path(folder_path, file)))
}
```

## Okay let's try it using Week 12 stuff:
```{r}
for (file in file_list) {
  # Extract the two-digit number from the file name
  file_number <- gsub("cbb|\\.csv", "", file)
  
  # I need to do something here that will pull each .csv based on file_number and then assign each value of file_number to a new data frame named "cbb*file_number*" 
  
  
}

```


## Data Exploration
```{r}
summary(cbb)
```

What kind of stuff do I have here?

- 4 character variables (TEAM, CONF, POSTSEASON, SEED)

- 20 numeric variables (G, W, ADJOE, ADJDE, BARTHAG, EFG_O, EFG_D, TOR, TORD, ORB, DRB, FTR, FTRD, X2P_O, X2P_D, X3P_O, X3P_D, ADJ_T, WAB, YEAR)

  - Very happy that there does not appear to be any NAs in the numeric stuff or in the character stuff 

summary(cbb13)

- This is only for 2013 so it doesn't have a year column but it has all the other columns. What kind of things can I glean from this data?


## Things I may need to do

- I'm gonna want to make postseason, possibly seed, possibly conference, and possibly team into factors with levels for easier plotting. 

- 


